Rough calculations

### 2015-12-03 notes/diffusion-local-times.Rmd
time scales

Let's compare the case where the cline is set up instantly or not:
```{r compare_rightness, cache=TRUE,dependson=c("fwds_back_pde","solve_pde_setup")}
layout(t(1:2))
fb.p <- clineplot(rev.soln,main="forwards-backwards")
st.p <- clineplot(pde.soln,main="stationary cline")
layout(1)
plot(fb.p,st.p,asp=1,pch=20,cex=0.5,xlab="forwards-backwards",ylab="stationary",col=rainbow(ncol(fb.p))[col(fb.p)])
abline(0,1,lwd=3)
```

@@

Let's check this agrees with the solution for the PDE.
According to Knight 1981 (Thm 7.4.3), $\zeta'(0) = - 2 r \zeta(0)$;
ReacTran lets you specify the 'flux' at the boundary is proportional to $\zeta(0)$,
where the flux is in our case $\sigma^2/2 \zeta'$.
First note that the solution depends on where we put the boundaries:
```{r reflected_pde_boundaries,cache=TRUE}
# NOTE: must install ReacTran from R-forge, e.g. 
#   install.packages("ReacTran",repos="http://r-forge.r-project.org")
# or else
source("tran.1D.R")
# to fix bug in using convective boundary conditions.
x.eps <- 5*mean(thegrid$dx)
solve.reflected <- function (x.up=-20) {
    thegrid <- setup.grid.1D(x.up=x.up, x.down=0, N = 5000)
    hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
    yinit <- rep(1.0,thegrid$N)
    tau <- 40
    tt <- seq(0,tau,length.out=20)
    r <- 1
    pde.fn <- function (t,y,parms,...) {
        # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
        tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=4*r)$dC
        list( tran - r * hetfn$mid * y )
    }
    soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
    attr(soln,"x") <- thegrid$x.mid
    return(soln)
}
pde.soln.0 <- solve.reflected(x.up=-5)
pde.soln.1 <- solve.reflected(x.up=-8)
pde.soln.2 <- solve.reflected(x.up=-20)
clineplot(pde.soln.0,x=attr(pde.soln.0,"x"),xlim=c(-5,0))
clineplot(pde.soln.1,x=attr(pde.soln.1,"x"),add=TRUE)
clineplot(pde.soln.2,x=attr(pde.soln.2,"x"),add=TRUE)
```

And, the PDE solution doesn't agree with the simulations very well: 
clearly, we need a different boundary condition at zero.
This is broken in ReacTran; I've modified the function to fix this.
```{r reflected_pde,cache=TRUE}
thegrid <- setup.grid.1D(x.up=-20, x.down=0, N = 5000)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
yinit <- rep(1.0,thegrid$N)
tau <- 40
tt <- seq(0,tau,length.out=20)
r <- 1
pde.fn <- function (t,y,parms,...) {
    # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
    tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=2*r )$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
ps <- clineplot(pde.soln,times=seq_along(tt),xlim=c(-8,0),legend=FALSE)
matlines( thegrid$x.mid, outer( thegrid$x.mid, tt, function(x,t) { zeta.reflected(x/sqrt(t),r*t*x.eps) }), lty=3, col=rainbow(length(tt)) )
legend("bottomleft",lty=1:2,legend=c("PDE","theory"))
```
Still not right... hm.

@@

Above we simulated the clines deriving from this process.
```{r sim_pushed, cache=TRUE}
rvals <- exp(seq(log(.001),log(0.5),length.out=10))
pushfn <- function(x,s){ifelse(x>0,-4*sqrt(s),0)}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
xinit <- seq(-8,8,length.out=nwalks)
sim.rw <- sim_diffusion( pushfn, tau, nwalks, ntimes, s=s, xinit=xinit )
sim.wt <- weighted_time( sim.rw, function (x) { x>0 } )
# distribution of time above zero depending on starting point
plot(xinit,sim.wt,xlab="initial position", ylab="time above zero")
```
We can also get the predicted clines by solving the PDE.
Note that here we are approximating $p(x) = \min(1,\exp(-4x\sqrt{s}))$.
```{r pushed_pde,cached=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
clinefn <- setup.prop.1D(grid=thegrid, func=function(x,s){ifelse(x>0,exp(-4*x*sqrt(s)),1.0)}, s=s)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){ifelse(x>0,1,0)})
yinit <- rep(1.0,thegrid$N)
tt <- seq(0,tau,length.out=100)
pde.solns <- lapply( rvals, function (r) {
        pde.fn <- function (t,y,parms,...) {
            tran <- tran.1D(C=y, A=clinefn, D=1/2, flux.up=0, flux.down=0, dx=thegrid)$dC
            if (any(is.na(tran))) { browser() }
            list( tran - r * hetfn$mid * y )
        }
        ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
      } )
layout(t(1:2))
for (k in c(6,10)) {
    clineplot(pde.solns[[k]],main=sprintf("r=%0.2f",tau,rvals[k]))
}
```

>>>>

Let's compare the case where the cline is set up instantly or not:
```{r compare_rightness, cache=TRUE,dependson=c("fwds_back_pde","solve_pde_setup")}
layout(t(1:2))
fb.p <- clineplot(rev.soln,thegrid,main="forwards-backwards")
st.p <- clineplot(pde.soln,thegrid,main="stationary cline")
layout(1)
plot(fb.p,st.p,asp=1,pch=20,cex=0.5,xlab="forwards-backwards",ylab="stationary",col=rainbow(ncol(fb.p))[col(fb.p)])
abline(0,1,lwd=3)
```

@@

Let's check this agrees with the solution for the PDE.
According to Knight 1981 (Thm 7.4.3), $\zeta'(0) = - 2 r \zeta(0)$;
ReacTran lets you specify the 'flux' at the boundary is proportional to $\zeta(0)$,
where the flux is in our case $\sigma^2/2 \zeta'$.
First note that the solution depends on where we put the boundaries:
```{r reflected_pde_boundaries,cache=TRUE}
# NOTE: must install ReacTran from R-forge, e.g. 
#   install.packages("ReacTran",repos="http://r-forge.r-project.org")
# or else
source("tran.1D.R")
# to fix bug in using convective boundary conditions.
x.eps <- 5*mean(thegrid$dx)
solve.reflected <- function (x.up=-20) {
    thegrid <- setup.grid.1D(x.up=x.up, x.down=0, N = 5000)
    hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
    yinit <- rep(1.0,thegrid$N)
    tau <- 40
    tt <- seq(0,tau,length.out=20)
    r <- 1
    pde.fn <- function (t,y,parms,...) {
        # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
        tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=4*r)$dC
        list( tran - r * hetfn$mid * y )
    }
    soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
    attr(soln,"x") <- thegrid$x.mid
    return(soln)
}
pde.soln.0 <- solve.reflected(x.up=-5)
pde.soln.1 <- solve.reflected(x.up=-8)
pde.soln.2 <- solve.reflected(x.up=-20)
clineplot(pde.soln.0,thegrid,x=attr(pde.soln.0,"x"),xlim=c(-5,0))
clineplot(pde.soln.1,thegrid,x=attr(pde.soln.1,"x"),add=TRUE)
clineplot(pde.soln.2,thegrid,x=attr(pde.soln.2,"x"),add=TRUE)
```

And, the PDE solution doesn't agree with the simulations very well: 
clearly, we need a different boundary condition at zero.
This is broken in ReacTran; I've modified the function to fix this.
```{r reflected_pde,cache=TRUE}
thegrid <- setup.grid.1D(x.up=-20, x.down=0, N = 5000)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
yinit <- rep(1.0,thegrid$N)
tau <- 40
tt <- seq(0,tau,length.out=20)
r <- 1
pde.fn <- function (t,y,parms,...) {
    # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
    tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=2*r )$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
ps <- clineplot(pde.soln,thegrid,times=seq_along(tt),xlim=c(-8,0),legend=FALSE)
matlines( thegrid$x.mid, outer( thegrid$x.mid, tt, function(x,t) { zeta.reflected(x/sqrt(t),r*t*x.eps) }), lty=3, col=rainbow(length(tt)) )
legend("bottomleft",lty=1:2,legend=c("PDE","theory"))
```
Still not right... hm.

@@

Above we simulated the clines deriving from this process.
```{r sim_pushed, cache=TRUE}
rvals <- exp(seq(log(.001),log(0.5),length.out=10))
pushfn <- function(x,s){ifelse(x>0,-4*sqrt(s),0)}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
xinit <- seq(-8,8,length.out=nwalks)
sim.rw <- sim_diffusion( pushfn, tau, nwalks, ntimes, s=s, xinit=xinit )
sim.wt <- weighted_time( sim.rw, function (x) { x>0 } )
# distribution of time above zero depending on starting point
plot(xinit,sim.wt,xlab="initial position", ylab="time above zero")
```
We can also get the predicted clines by solving the PDE.
Note that here we are approximating $p(x) = \min(1,\exp(-4x\sqrt{s}))$.
```{r pushed_pde,cached=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
clinefn <- setup.prop.1D(grid=thegrid, func=function(x,s){ifelse(x>0,exp(-4*x*sqrt(s)),1.0)}, s=s)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){ifelse(x>0,1,0)})
yinit <- rep(1.0,thegrid$N)
tt <- seq(0,tau,length.out=100)
pde.solns <- lapply( rvals, function (r) {
        pde.fn <- function (t,y,parms,...) {
            tran <- tran.1D(C=y, A=clinefn, D=1/2, flux.up=0, flux.down=0, dx=thegrid)$dC
            if (any(is.na(tran))) { browser() }
            list( tran - r * hetfn$mid * y )
        }
        ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
      } )
layout(t(1:2))
for (k in c(6,10)) {
    clineplot(pde.solns[[k]],thegrid,main=sprintf("r=%0.2f",tau,rvals[k]))
}
```

### 2015-12-03 notes/rough-calculations.Rmd
time scales

    ```{r forwards_selected_cline,cache=TRUE}
    thegrid <- setup.grid.1D(x.up=-100, x.down=100, N = 5000)
    svals <- exp(seq(log(1e-4),log(0.5),length.out=6))
    fwds.solns <- lapply(svals, function(s) {
            tt <- seq(0,3/s,length.out=40)
            forwards_pde(s=s,times=tt,grid=thegrid) 
        } )
    for (k in seq_along(svals)) {
        clineplot( fwds.solns[[k]], main=sprintf("s=%f",svals[k]), col=ifelse( fwds.solns[[k]][,1]>1/svals[k], adjustcolor("black",0.5), adjustcolor("red",0.5) ) )
        abline(v=c(-1,1)/sqrt(svals[k]),lty=3,col='green',lwd=2)
    }
    ```

@@

a.  The haplotype of a $B$ allele at distance $x$ into the $A$ side should be of length about $\sigma/(sx)$, as long as this is less than $1/\tau$. (this is fairly restrictive)

@@

a.  This is (probably) the Laplace transform of the time since the $B$ allele
    last had an ancestor on the $B$ side,
    i.e., for a $B$ lineage to make it back to the $B$ side.
    Since lineages on the wrong side move at speed $\sigma/s$,
    if the allele is at position $x$, 
    this will take about $sx/\sigma$;
    and so the haplotype should be about $\sigma/(sx)$.
    This gives the size of the chunk carried with it on this excursion;
    this will be of the same type with high probability at least if the length is less than $1/\tau$,
    as above.

>>>>

    ```{r forwards_selected_cline,cache=TRUE}
    thegrid <- setup.grid.1D(x.up=-100, x.down=100, N = 5000)
    svals <- exp(seq(log(1e-4),log(0.5),length.out=6))
    fwds.solns <- lapply(svals, function(s) {
            tt <- seq(0,3/s,length.out=40)
            forwards_pde(s=s,times=tt,grid=thegrid) 
        } )
    for (k in seq_along(svals)) {
        clineplot( fwds.solns[[k]], thegrid, main=sprintf("s=%f",svals[k]), col=ifelse( fwds.solns[[k]][,1]>1/svals[k], adjustcolor("black",0.5), adjustcolor("red",0.5) ) )
        abline(v=c(-1,1)/sqrt(svals[k]),lty=3,col='green',lwd=2)
    }
    ```

@@

a.  The haplotype of a $B$ allele at distance $x$ into the $A$ side should be of length about $s \sigma/x$, as long as this is less than $1/\tau$. (this is fairly restrictive)

@@

a.  This is (probably) the Laplace transform of the time since the $B$ allele
    last had an ancestor on the $B$ side,
    i.e., for a $B$ lineage to make it back to the $B$ side.
    Since lineages on the wrong side move at speed $\sigma s$,
    if the allele is at position $x$, 
    this will take about $x/(s\sigma)$;
    and so the haplotype should be about $s\sigma/x$.
    This gives the size of the chunk carried with it on this excursion;
    this will be of the same type with high probability at least if the length is less than $1/\tau$,
    as above.
